\LARGE{{\bf{Gravitational Wave Signals from Multiple Hidden Sectors} \\ }}

### 2019-09-25 gwnnDraft02.tex

The recent experimental detection of gravitational waves~\cite{Abbott:2016blz} gives humanity a new way to observe the universe. Future experiments~\cite{Seto:2001qf,Crowder:2005nr,Harry:2006fi,Sathyaprakash:2012jk,Janssen:2014dka,2017arXiv170200786A,Sato:2017dkf} will greatly expand the frequency range observable. Thus far, experiments have only observed recent events such as black hole mergers, but phase transitions in the early universe can leave an imprint as a \textit{stochastic} gravitational wave background~\cite{Witten:1984rs,Hogan:1984hx,Hogan:1986qda,PhysRevLett.65.3080,Caprini:2015zlo,Mazumdar:2018dfl}. Thus searches for this background of gravitational waves can give direct information of the history of the universe before big bang nucleosynthesis. 

@@

In cosmological terms, we can envision a sector's thermal history unfolding where as the plasma cools below each quark mass threshold, said quarks are frozen out. At a certain point, the sector arrives at the QCD phase transition and confinement occurs (provided there are light enough quarks to confine) --- if this occurs when $\geq 3$ quarks are at a much lower scale or all quarks have already frozen out, we get the desired phase transition.   

@@

However, if we move away from the original $N$naturalness reheating mechanism and begin exploring mirror sectors with large vevs and with relative energy densities $\rho_i / \rho_{SM} \sim 10\%$, a possibility allowed by current constraints, we can have sectors with relatively high dark QCD scales that produce detectable gravitational waves. From Eq.~(\ref{eqn:lambda2}) we can determine the confinement scale of an arbitrary mirror sector. If we take Higgs vevs as high as the GUT scale $\sim 10^{16}$ GeV, then we can gave confinement scales as high as $\sim 38 \,\gev$.  The signals of this sector and other test cases like it are explored in Sec. \ref{sec:gw}. 

@@

In general, the multi-hidden sector models explored here feature feature of huge number of (nearly) massless degrees of freedom. Dark photons and dark neutrinos abound in these sectors and, assuming a relatively high reheat temperature, the leptons, quarks, and heavy bosons of these sectors can also be relativistic. In $N$naturalness this feature is realized quite dramatically: each of the $N$ of sectors possess relativistic degrees of freedom. The presence of these particles can have two main effects: extra relativistic particles can alter the expansion history of the universe through changes to the energy density or hidden sectors can feature annihilations that reheat the photons or neutrinos of our sector near Big Bang Nucleosynthesis (BBN) and affect the light element abundances. The effective number of neutrino species, $N_{eff}$, is impacted by these contributions and, as such, is the strictest constraint that must be dealt with when studying these type of multi-phase transition models. 

@@

\frac{\beta}{H} = T^{nuc}_h \frac{dS_{E}(T)}{dT}\bigg|_{T = T^{nuc}}.

@@

Due to the lack of an analytic potential, we take $\beta/H$ to be a standard value currently seen in the field~\cite{}. There has been a lot of work in recent years in the estimation of $\beta/H$ using first-order chiral effective theories and the Polyakov-Nambu-Jona-Lasinio (PNJL) models~\citep{Helmboldt:2019pan,Bai:2018dxf}. \draftnote{Are there other papers? One paper is not a lot of work}. Although, these studies claim a large range of values and thus a consensus has not yet been reached.

@@

\Omega_{\textrm{GW}} =  \sum_{i}^{N} \mathcal{A}^{i}\, \Omega_{\textrm{GW}}^{i,*}((a_{0}/a)_{i} f) 

@@

It is safe to assume that a subset of parameters of the SFOPT do not differ between sectors within $N$natrualness or other decoupled hidden sector models.\draftnote{what?} In the case of N-copies of dark QCD-like phase transitions, the relativistic degrees of freedom, phase transition rate,  the dark QCD scale, are all similar. This makes the redshifting factor, $\mathcal{A}^i$, independent of sector number.

@@

Gravitational wave signals continue to grow in importance as phenomenological si
We can think of these results providing 
This demonstrates the power of GW signals to probe deep into the unknown arena of complex hidden sectors --- whether there are 1, 5, or more hidden sectors 
Further, should a stochastic GW background be detected, deviations from a standard power law curve would indicate a more complicated hidden sector.  
Questions, of course, still remain: how can the estimations of $\alpha$ and $\beta \ H$ be improved?
%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\bibliographystyle{apsrev4-1}
\bibliography{refs}
\end{document}

>>>>

The recent experimental detection of gravitational waves~\cite{Abbott:2016blz} gives humanity a new way to observe the universe. Future experiments~\cite{Seto:2001qf,Crowder:2005nr,Harry:2006fi,Sathyaprakash:2012jk,Janssen:2014dka,2017arXiv170200786A,Sato:2017dkf} will greatly expand the frequency range observable. Thus far, experiments have only observed recent events such as black hole mergers, but phase transitions in the early universe can leave an imprint as a \textit{stochastic} gravitational wave background~\cite{Witten:1984rs,Hogan:1984hx,Hogan:1986qda,PhysRevLett.65.3080,Caprini:2015zlo,Mazumdar:2018dfl}. Thus, searches for this background of gravitational waves can give direct information of the history of the universe before big bang nucleosynthesis. 

@@

In cosmological terms, we can envision a sector's thermal history unfolding, whereas the plasma cools below each quark mass threshold and said quarks are frozen out. At a certain point, the sector arrives at the QCD phase transition and confinement occurs (provided there are light enough quarks to confine) --- if this occurs when $\geq 3$ quarks are at a much lower scale or all quarks have already frozen out, we get the desired phase transition.   

@@

However, if we move away from the original $N$naturalness reheating mechanism and begin exploring mirror sectors with large vevs and with relative energy densities $\rho_i / \rho_{SM} \sim 10\%$, a possibility allowed by current constraints, we can have sectors with relatively high dark QCD scales that produce detectable gravitational waves. From Eq.~(\ref{eqn:lambda2}) we can determine the confinement scale of an arbitrary mirror sector. If we take Higgs vevs as high as the GUT scale $\sim 10^{16}$ GeV, then we can get confinement scales as high as $\sim 38 \,\gev$.  The signals of this sector and other test cases like it are explored in Sec. \ref{sec:gw}. 

@@

In general, the multi-hidden sector models explored feature a huge number of (nearly) massless degrees of freedom. Dark photons and dark neutrinos abound in these sectors and, assuming a relatively high reheat temperature, the leptons, quarks, and heavy bosons of these sectors can also be relativistic. In $N$naturalness this feature is realized quite dramatically: each of the $N$ of sectors possess relativistic degrees of freedom. The presence of these particles can have two main effects: extra relativistic particles can alter the expansion history of the universe through changes to the energy density or hidden sectors can feature annihilations that reheat the photons or neutrinos of our sector near Big Bang Nucleosynthesis (BBN) and affect the light element abundances. The effective number of neutrino species, $N_{eff}$, is impacted by these contributions and, as such, is the strictest constraint that must be dealt with when studying these type of multi-phase transition models. 

@@

\begin{equation}\label{eqn:BH}
\frac{\beta}{H} = T^{nuc}_h \frac{dS_{E}(T)}{dT}\bigg|_{T = T^{nuc}_h}.

@@

Due to the lack of a general analytic QCD potential, it is not possible to use Eq.~(\ref{eqn:BH}) to calculate $\beta/H$. There are dimensional arguments~\cite{Hogan:1984hx,Hogan:1986qda} which predict $\beta/H \sim 4 \textrm{Log}(M_{p}/ \Lambda_{QCD,i})$, although these arguments are riddled with assumptions on the potential. In more recent work, some authors~\citep{Helmboldt:2019pan,Bai:2018dxf} have attempted to estimate it using first-order chiral effective theories and the Polyakov-Nambu-Jona-Lasinio (PNJL) models. These studies claim a large range of values and thus a consensus has not yet been reached.  Under these circumstances, we take the optimistic case in which $\beta/H$ is of order $\mathcal{O}(10)$.

@@

\Omega_{\textrm{GW}} =  \sum_{i}^{N} \mathcal{A}^{i}\, \Omega_{\textrm{GW}}^{i,*}((a_{0}/a)_{i} f) .

@@

We assume that the parameters of the SFOPT do not differ between sectors: the relativistic degrees of freedom, phase transition rate, and the dark QCD scale, are all similar. This makes the redshifting factor $\mathcal{A}^i$ independent of sector number.

@@

As detection capabilities increase, gravitational wave signals continue to grow in importance as phenomenological signatures that can offer us a unique glimpse into the universe as it was in the early epochs. The space-based interferometers planned for the next generation of GW experiments will be sensitive enough to begin searching for signals of the cataclysmic disruption of space-time due to SFOPT. As we inch closer to these measurements becoming available, it becomes important to develop ways to analyze and understand this data.

Here, we examined several scenarios (including $N$naturalness) that involve multiple hidden sectors and calculated the GW profiles present. Our GW projections unambiguously demonstrate that scenarios with multiple hidden sector SFOPTs can begin to be probed by next generation experiments. Further, our results provide a framework for understanding and using GW signals in two opposite ways: first as a unique signal for specific theories featuring multiple SFOPTs and also as a way to explain signals that deviate from a standard power law spectrum.

In the former case, this demonstrates the power of GW signals to probe deep into the unknown arena of complex hidden sectors. Whether a model predicts of the presence of 1, 5, or more hidden sectors, any SFOPT can potentially contribute a GW signal and the sum of these signals can create a unique GW background; ultimately providing another means to test and constrain theories.

Shifting to the latter scenario, should a stochastic GW background be detected, deviations from a standard power law curve would indicate a more complicated hidden sector.  Explicitly, a broadening or distortion of the signal around the peak frequency could point to a multi-SFOPT scenario --- and gently guide us in the direction of multiple hidden sectors.

%Questions, of course, still abound: how can the %estimations of $\alpha$ and $\beta/H$ be improved?

%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{apsrev4-1}
\bibliography{refs}
\end{document}
